---
title: "Homework Class 17: Downstream Analysis"
author: Kiana Bohanon (A17802316)
format: pdf
toc: TRUE
---

## Downstream Analysis

We can use R and Bioconductor tools to further explore this large scale data-set. 
```{r}
library(tximport)

# setup the folder and file-names to read
folders <- dir(pattern="SRR21568*")
samples <- sub("_quant", "", folders)
files <- file.path( folders, "abundance.h5" )
names(files) <- samples

txi.kallisto <- tximport(files, type = "kallisto", txOut = TRUE)
```
```{r}
head(txi.kallisto$counts)
```
This is our estimated transcript counts for each sample in R.

To see how many transcipts we have for each sample:
```{r}
colSums(txi.kallisto$counts)
```

And how many transcripts are detected in at least one sample:
```{r}
sum(rowSums(txi.kallisto$counts)>0)
```

Filtering out the annotated transcripts with no reads:
```{r}
to.keep <- rowSums(txi.kallisto$counts) > 0
kset.nonzero <- txi.kallisto$counts[to.keep,]
```
And those with no change over the samples:
```{r}
keep2 <- apply(kset.nonzero,1,sd)>0
x <- kset.nonzero[keep2,]
```


## Principal Component Analysis

We can now apply any exploratory analysis technique to this counts matrix. As an example, we will perform a PCA of the transcriptomic profiles of these samples.


```{r}
pca <- prcomp(t(x), scale=TRUE)
summary(pca)
```

Now we can use the first two principal components as a co-ordinate system for visualizing the summarized transcriptomic profiles of each sample:
```{r}
plot(pca$x[,1], pca$x[,2],
     col=c("blue","blue","red","red"),
     xlab="PC1", ylab="PC2", pch=16)
```

> Q. Use ggplot to make a similar figure of PC1 vs PC2 and a separate figure PC1 vs PC3 and PC2 vs PC3.

```{r}
library(ggplot2)
library(ggrepel)

mycols <- c("blue","blue","red","red")

ggplot(pca$x) +
  aes(PC1, PC2, label=rownames(pca$x)) +
  geom_point( col=mycols ) +
  geom_text_repel( col=mycols ) +
  theme_bw()
```

This ggplot makes it clear that PC1 separates the control samples (SRR2156848 and SRR2156849) from the two enhancer-targeting samples (SRR2156850 and SRR2156851), implying that there are significant differences between the treated and control samples. 

## Differential-expression Analysis

We can use DESeq2 to complete the differential-expression analysis that we are already familiar with:

```{r}
library(DESeq2)

sampleTable <- data.frame(condition = factor(rep(c("control", "treatment"), each = 2)))
rownames(sampleTable) <- colnames(txi.kallisto$counts)

dds <- DESeqDataSetFromTximport(txi.kallisto,
                                sampleTable, 
                                ~condition)
```
```{r}
dds <- DESeq(dds)

res <- results(dds)
head(res)
```

