---
title: "Class 9: Candy Mini Project"
author: "Kiana Bohanon (PID: A17802316)"
format: pdf
toc: true
---
## Background

Our task is to explore the candy dataset to find out answers to various questions about the datset.

## Importing data

Importing candy data by downloading the csv file, placing it in the project directory, and then using the `read.csv()` code to check it was inputted properly. 

```{r}
#read.csv("candy-data.csv")
candy_file <- "candy-data.csv"
# Next, fix the column name to match the one on the instructions by treating the first column as row names
candy = read.csv(candy_file, row.names=1)
head(candy)
```

## What is in the dataset?

> Q1. How many different candy types are in this dataset?

There are `r nrow(candy)` rows in this dataset. 

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```
There are 38 fruity candy types in the dataset.

## What is your favorite candy?

`winpercent()` is the percentage of people who prefer a candy over another randomly chosen candy from the dataset. Higher values indicate a more popular candy. 

For example:
```{r}
candy["Twix",]$winpercent
```

> Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

My favorite candy is Sour Patch Kids, and it's winpercent value is 59.864.
```{r}
candy["Sour Patch Kids",]$winpercent
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```
The winpercent value for Kit Kat is 76.7686.

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```
The winpercent value for Tootsie Roll Snack Bars is 49.6535.

**Side-note** the skim() function
This function can give you a quick overview of a given dataset. First install the package. 
```{r}
library("skimr")
# skim(candy)
```

From your use of the skim() function use the output to answer the following:

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes, winpercent is out of 100 instead of 1 like the other variables. 

> Q7. What do you think a zero and one represent for the candy$chocolate column?

I think a zero and a one represent that the candy$chocolate column is totally complete and not missing anything. 

## Exploratory analysis

> Q8. Plot a histogram of winpercent values

Base R:
```{r}
hist(candy$winpercent)
```
ggplot:
```{r}
library(ggplot2)
ggplot(candy) + aes(winpercent) + geom_histogram(binwidth=10)
```

> Q9. Is the distribution of winpercent values symmetrical?

No, it is skewed to the right. 

> Q10. Is the center of the distribution above or below 50%?

```{r}
summary(candy$winpercent)
```

The center of the distribution is below 50%, as the median is 47.83. 

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

1. Find all chocolate candy.
2. Get their winpercent values and find the mean.
3. Find all fruity candy.
4. Get their winpercent values and find the mean.
5. Compare the two.

```{r}
choc.win <- candy$winpercent[candy$chocolate == 1]
mean(choc.win)
fruit.win <- candy$winpercent[candy$fruity == 1]
mean(fruit.win)
```
On average, chocolate candy is higher ranked. 

> Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fruit.win)
```

## Overall Candy Rankings

> Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent),], n=5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
tail(candy[order(candy$winpercent),], n=5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)
ggplot(candy) + aes(winpercent, rownames(candy)) + geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
library(ggplot2)
ggplot(candy) + aes(winpercent, reorder(rownames(candy),winpercent)) + geom_col()
```

## Adding Color

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 

```
Now, for the first time, using this plot we can answer questions like:

> Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy is Sixlets.

> Q18. What is the best ranked fruity candy?

The best ranked fruity candy is Starburst.

## Taking a look at pricepercent

We can examine the value for money, or what the best candy is for the least amount of money. 

```{r}
library(ggrepel)

# How about a plot of win vs price
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```
> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Miniatures

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```
Of these, the least popular is Nik L Nip. 

## Exploring the correlation structure

We can use correlation to view the results with the **corrplot** package to plot a correlation matrix. 
```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```
> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Examining this plot, two variables that are anti-correlated are fruity and chocolate.

> Q23. Similarly, what two variables are most positively correlated?

Two variables that are most positively correlated are winpercent and chocolate. 

## Principal Component Analysis

Let's apply PCA using the `prcomp()` function to our candy dataset.
```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```
Now we can plot our main PCA score plot of PC1 vs. PC2.
```{r}
plot(pca$x[,1:2])
```
We can change the plotting character and add some color:
```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```
> ggplot works best when you supply an input data.frame that includes a separate coloumn for each of the aesthetics you would like displayed in your final plot. To do this, we make a new data.frame that contains our PCA results with all the rest of our candy data. 

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)
p
```
We can use the ggrepel package and the function `ggrepel::geom_text_repel()` to label the plot. 
```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```
You can also generate an interactive plot that you can mouse over to see labels:
```{r}
#library(plotly)
#ggplotly(p)
```

Take a quick look at PCA our loadings.
```{r}
ggplot(pca$rotation) +
  aes(x=PC1, y=reorder(rownames(pca$rotation), PC1)) +
  geom_col()
```
> Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

The variables that are picked up strongly by PC1 in the positive direction are fruity, pluribus, and hard. This makes sense to me because fruity candy is usually hard and comes in multiples (i.e. Skittles and Starburst). We saw this relationship highlighted previously in the correlation figure from the corrplot package. 

## Summary

> Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

Based on these analyses, a "winning" candy tends to be a combination of chocolate and bar, but not fruity. The visualizations in histograms and bar plots show that candy with chocolate tends to have a high winpercent. The correlation analysis also supports this by showing positive correlation between winpercent and chocolate. The PCA also supports this by showing that the characteristics of chocolate and bar are in the same direction and strength, indicating they are correlated features that contribute positively to overall candy preference.