---
title: "Class 12 Lab"
author: "Kiana Bohanon (A17802316)"
format: pdf
toc: TRUE
---

## Section 1. Proportion of G/G in a Population

Downloaded a CSV file from Ensemble < https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39835095-39955096;v=rs8067378;vdb=variation;vf=959672880;sample= >

Here we read this CSV file:
```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```
```{r}
table(mxl$Genotype..forward.strand.) /nrow(mxl)
```

## Section 4: Population Scale Analysis [HOMEWORK]

One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. So, you processed about ~230 samples and did the normalization on a genome level. Now, you want to find whether there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3 expression.

How many samples do we have?

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes. 

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

Sample size for each genotype: 
```{r}
table(expr$geno)
```
There are 108 A/A, 233 A/G, and 121 G/G.

Median expression levels for each genotype:
```{r}
tapply(expr$exp, expr$geno, median, na.rm=TRUE)
```
For A/A, the median expression level is 31.24. For A/G, it is 25. For G/G, it is 20. 

Lets make a boxplot:

```{r}
library(ggplot2)
```


> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3? Hint: An example boxplot is provided overleaf – yours does not need to be as polished as this one.

```{r}
ggplot(expr) + aes(x=geno, y=exp, fill=geno) +geom_boxplot(notch=TRUE)
```

Based on this plot, you can infer that there is a difference between the A/A and G/G relative expression value. So the SNP does effect the expression of ORMDL3, because if you have the G/G genotype, you are more likely to see reduced expression. 
