---
title: "Class 7 Machine Learning 1"
author: "Kiana Bohanon (PID: A17802316)"
format: pdf
toc: true
---

## Background

Today we will begin our exploration of some important machine learning methods with a focus on **clustering** and **dimensionality reduction**.

To start testing these methods, let's make up some sample data to cluster where we know what the answer should be. 

```{r}
hist(rnorm(3000,mean=10,sd=))
```

> Q. Can you generate 30 numbers centered at +3 and 30 numbers at -3 taken at random from a normal distribution?

```{r}
tmp <- c(rnorm(30, mean=3), rnorm(30,mean=-3))

x <- cbind(x=tmp, y=rev(tmp))

plot(x)

```

## K-means Clustering

The main function in "base R" for K-means clustering is called `kmeans()`, let's try it out:
```{r}
k <- kmeans(x, centers=2)
k
```
> Q. What component of your kmeans result object has the cluster centers?

```{r}
k$centers
```

> Q. What component of your kmeans result object has the cluster size (i.e. how many points are in each cluster)?

```{r}
k$size
```

> Q. What component of your kmeans result object has the cluster membership vector (i.e. the main clustering result: which points are in which clusters)?

```{r}
k$cluster
```

> Q. Plot the results of clustering (i.e. our data colored by the clustering result) along with the cluster centers.

```{r}
plot(x, col=k$cluster)
points(k$centers, col="blue", pch=15, cex=2)
```

> Q. Can you run `kmeans()` again and cluster `x` into 4 clusters, and plot the results just like we did above with coloring by cluster and the cluster centers shown in blue?

```{r}
k4 <- kmeans(x,centers=4)
plot(x, col=k4$cluster)
points(k4$centers, col="blue", pch=15, cex=2)

```

> **Key Point:** Kmeans will always return the clustering that we ask for (this is the "K" or "centers" in Kmeans)!

```{r}
k$tot.withinss
```

## Hierarchical Clustering

The main function for hierarchical clustering in base R is called `hclust()`. One of the main differences with respect to the `kmeans()` function is that you can not just pass your input data directly to `hclust()` - it needs a "distance matrix". We can get this from lots of places, including the `dist()` function. 

```{r}
d <- dist(x)
hc <- hclust(d)
plot(hc)
```

We can "cut" the dendrogram or "tree" at a given height to yield our "clusters". For this we use the function `cutree()`
```{r}
plot(hc)
abline(h=10,col="red")
grps <- cutree(hc, h=10)
grps
```
```{r}
grps
```

> Q. Plot our data `x` colored by the clustering result from `hclust()` and `cutree()`?

```{r}
grps <- cutree(hc, h=10)
plot(x, col=grps)
```

## Principal Component Analysis (PCA)

PCA is a popular dimensionality reduction technique that is widely used in bioinformatics. 

## PCA of UK food data

Read data on food consumption in the UK:
```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
x
```
It looks like the row names are not set properly. We can fix this.
```{r}
rownames(x) <- x[,1]
x <- x[,-1]
x
```
A better way to do this is to fix the row names assignment at import time
```{r}
x <- read.csv(url, row.names=1)
x
```

> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
dim(x)
```

> Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

I preferred the second approach best because the code was a lot simpler and it was more robust.

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

> Q3: Changing what optional argument in the above barplot() function results in the following plot?
Changing the beside argument to FALSE.

```{r}
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))
```

> Q4: ggplot part skipped per instructions

## Pairs plots and heatmaps

> Q5: We can use the pairs() function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)
```
In this plot, the x and y axis coordinate to the different countries, and we are examining the data against each other. If the countries have the exact same consumption value, the points would lie exactly on the diagonal. If the points are off of the diagonal, then they have different consumption vales. 

## Heatmap
We can install the **pheatmap** package with the `install.packages()` command that we used previously. Remember that we always run this in the console and not a code chunk1 in our quarto document. 

```{r}
library(pheatmap)

pheatmap( as.matrix(x) )
```
Of all these plots, really only the `pairs()` plot was useful. This took a bit to interpret and will only scale when I am looking at bigger datasets.

> Q6. Based on the pairs and heatmap figures, which countries cluster together and what does this suggest about their food consumption patterns? Can you easily tell what the main differences between N. Ireland and the other countries of the UK in terms of this data-set?

Based on the pairs and heatmap figures, you can sort of tell that England, Wales, and Scotland tend to the cluster together and that their food consumption patterns are similar. You can tell that N. Ireland has the most differences in fruit, meat, and potatoes compared to the other countries. 

## PCA to the rescue

The main function in "base R" for PCA is called `prcomp()`.

```{r}
pca <- prcomp(t(x))
summary(pca)
```

> Q. How much variance is captured in the first PC?

67.44%
> Q. How many PCs do I need to capture at least 90% of the total vairance in the dataset?

Two PCs capture 96.5% of the total variance.

> Q. Plot our main PCA result. Folks can call this different things depending on their field of study e.g. "PC Plot", "ordienation plot", "score plot"...

```{r}
attributes(pca)
```

To generate our PCA score plot we want the `pca$x` component of the result object.
```{r}
pca$x
```

```{r}
plot(pca$x[,1],pca$x[,2])
```
```{r}
my_cols <- c("orange","red","blue","darkgreen")
plot(pca$x[,1],pca$x[,2], col=my_cols,pch=16)
```
```{r}
library(ggplot2)
ggplot(pca$x) + aes(PC1, PC2) + geom_point(col=my_cols)
```

## Digging deeper (variable loading)

How do the original variables (i.e. the 17 different foods) contribute to our new PCs?

```{r}
ggplot(pca$rotation) +
  aes(x = PC1, 
      y = reorder(rownames(pca$rotation), PC1)) +
  geom_col(fill = "steelblue") +
  xlab("PC1 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```


